\documentclass[a4paper]{article}

\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{listings}
\usepackage{color}
\usepackage[nounderscore]{syntax}

% Programming language

\newcommand{\pfunction}[3]{\mathtt{function} \ #1(#2) \ #3}
\newcommand{\psequence}[2]{#1 ; #2}
\newcommand{\pifelse}[3]{\mathtt{if} \ (#1) \ #2 \ \mathtt{else} \ #3}
\newcommand{\pwhile}[2]{\mathtt{while} \ (#1) \ #2}
\newcommand{\pdowhile}[2]{\mathtt{do} \ #1 \ \mathtt{while} \ (#2)}
\newcommand{\plassign}[2]{#1 := #2}
\newcommand{\passign}[2]{[#1] := #2}
\newcommand{\pderef}[2]{#1 := [#2]}
\newcommand{\pfuncall}[3]{#1 := #2(#3)}
\newcommand{\pcas}[4]{#1 := \mathtt{CAS}(#2, #3, #4)}
\newcommand{\palloc}[2]{#1 := \mathtt{alloc}(#2)}
\newcommand{\preturn}[1]{\mathtt{return} \ #1}
\newcommand{\pskip}[0]{\mathtt{skip}}
\newcommand{\pfork}[2]{\mathtt{fork} \ #1(#2)} 

\newtheorem*{extension}{Extension}


\lstdefinelanguage{caper}
  {morekeywords={region,guards,interpretation,actions,function,requires,ensures,var,alloc,return,CAS,do,while,if,else},
  sensitive=true,
  morecomment=[l]{//},
  morecomment=[s]{/*}{*/}
  }
\lstset{
  basicstyle=\small,
  keywordstyle=\color{red}
}

\begin{document}
\title{\textsc{Caper}: Design Document}
\author{}

\maketitle

This document specifies \textsc{Caper}, a prototype tool for (semi-)automatic program verification using Concurrent Abstract Predicates (CAP).
Programs will be annotated with pre- and postconditions on functions, loop invariants and other assertions, definitions or proof hints.
\textsc{Caper} will attempt to verify programs against these specifications.

\section{Programming Language}

The \textsc{Caper} programming language has:
\begin{itemize}
  \item functions (procedures) as the basic top-level elements;
  \item function-scoped mutable variables;
  \item a shared heap, mutable with primitive atomic operations;
  \item integer arithmetic;
  \item conditional and loop constructs;
  \item fork/join concurrency.
\end{itemize}

The language 

\section{Program Logic}

The \textsc{Caper} program logic is based on CAP and inspired by iCap and TaDA.
As in CAP, assertions describe resources owned by the local thread and resources belonging to shared regions.
Shared regions are subject to protocols, which determine how their resources may be used.

As in iCap, but unlike in CAP, the state of shared regions are abstracted.
That is, instead of directly specifying the resources currently belonging to the region, an assertion specifies the possible abstract states of a region, which are associated with concrete resources by an interpretation function.
Abstract states of regions are integers.

\begin{extension}
  Permit other datatypes as the abstract states of regions.
\end{extension}

Each region is associated with a protocol that determines how its (abstract) state may change.
In order for a thread to perform a state change, it must have access to a resource (called a guard) associated with the state change in the protocol.
In CAP, the guard is always a non-zero fractional permission to a single named (and parametrised) guard.
In TaDA, guards are elements of an arbitrary separation algebra associated with the region.
\textsc{Caper} takes a middle ground: guards are elements of a separation algebra of a certain form.

\subsection{Guards and Guard Types}

A guard type specifies the separation algebra of guards that is associated with a region.
Guard types are composed from primitive guard types.
Each primitive guard type is associated with a name, which must be unique within the overall guard type.
There are three kinds of primitive guard types:
\begin{itemize}
  \item Unique guards consist of a name and nothing else: there can be only one such guard for a region.
  \item Permission guards consist of a name and a permission value: there may be multiple permission guards for a region, whose permissions compose to give full permission.
  \item Parametrised guards consist of a name together with an integer value: there is one of these guards for each integer.
    The syntax for guards makes it possible to represent the combination of all parametrised guards of a given name, excluding a finite set.
\end{itemize}

\begin{extension}
  Permit combinations of parameters and permissions.
\end{extension}

There are two operators for combining guard types: product and sum.
The product of two guard types $a \times b$ can be thought of as the set of pairs of guards from $a$ and guards from $b$.
(This corresponds to a direct product on the separation algebras.)

The sum of two guard types $a + b$ contains all guards from $a$ and all guards from $b$.
(This can be thought of as the (disjoint/discriminated) union of guard types $a$ and $b$.)
However, we identify the ``full'' guards of $a$ and $b$.

The sum construction allows us to specify rich forms of capabilities, such as deny-guarantee or insert-remove-definite.

\subsection{Protocols}

Here are some examples of protocol specifications.

For a lock:
\[
\begin{array}{r@{\ :\ } c @{\ \rightsquigarrow\ } c}
  A [ \_ ] & 0 & 1 \\
  B & 1 & 0
\end{array}
\]
$A [ \_ ]$ indicates that we have some (unspecified but non-zero) permission on $A$.

\[
\begin{array}{c @{\ |\ } r@{\ :\ } c @{\ \rightsquigarrow\ } c}
  n > 0 & A (n) & n & -n \\
  n > 0 & B & -n & n+1 \\
  n > 0 & B * C & 0 & n \\
\end{array}
\]

Logical variables may occur in protocol specifications.
We can stipulate constraints on the logical variables as side-conditions.

\section{First Example}

We shall use a simple compare-and-swap lock as the motivating example for the \textsc{Caper} tool.
The code, together with annotations, is shown in Figure~\ref{fig:caslock}.

\begin{figure}
\begin{center}
\begin{lstlisting}[language=caper,mathescape,columns=flexible]
// Compare-and-swap lock

region Lock(r,x) {
  guards %LOCK * UNLOCK;
  interpretation {
    0 : x $\mapsto$ 0 * r@UNLOCK;
    1 : x $\mapsto$ 1;
  }
  actions {
    LOCK[_] : 0 $\rightsquigarrow$ 1;
    UNLOCK : 1 $\rightsquigarrow$ 0;
  }
}

function makeLock()
  requires emp;
  ensures Lock(r,ret,0) * r@LOCK[1]; {
    var v;
    v := alloc(1);
    [v] := 0;
    return v;
}

function lock(x)
  requires Lock(r,x,_) * r@LOCK[p];
  ensures Lock(r,x,1) * r@(LOCK[p] * UNLOCK); {
    var b;
    do {
        b := CAS(x, 0, 1);
    } while (b = 0);
}

function unlock(x)
  requires Lock(r,x,1) * r@UNLOCK;
  ensures Lock(r,x,_); {
    [x] := 0;
}
\end{lstlisting}
\end{center}
\caption{Compare-and-swap lock example.}
\label{fig:caslock}
\end{figure}

It may be necessary to add an invariant assertion to the loop.

\section{Grammar}

\[
\begin{array}{rcl}
  F & ::= & \pfunction{\mathtt{f}}{\mathtt{x}}{C} \\
  C & ::= & \psequence{C}{C} \ | \ \pifelse{B}{C}{C} \ | \ \pwhile{B}{C} \ | \ \pdowhile{C}{B} \ | \ \\
  & & \plassign{\mathtt{x}}{E} \ | \ \pderef{\mathtt{x}}{E} \ | \ \passign{E}{E} \ | \ \palloc{\mathtt{x}}{E} \ | \ \\
  & & \pcas{\mathtt{x}}{E}{E}{E} \ | \ \pfuncall{\mathtt{x}}{\mathtt{f}}{E} \ | \ \preturn{E} \ | \ \pskip \ | \ \pfork{\mathtt{f}}{E} \\
  E & ::= & \mathtt{x} \ | \ n \ | \ ( E ) \ | \ E + E \ | \ E - E \ | \ E * E \ | \ E / E\\
  B & ::= & \mathtt{true} \ | \ \mathtt{false} \ | \ ( B ) \ | \ E = E \ | \ E \ \text{!=} \ E \ | \ E > E \ | \ \\
  & & E \ge E \ | \ E < E \ | \ E \le E \ | \ E \ \mathtt{or} \ B \ | \ B \ \mathtt{and} \ B \ | \ \mathtt{not} \ B
\end{array}
\]

\begin{grammar}
  <lower> ::= "a" | "b" | \ldots | "z"

  <upper> ::= "A" | "B" | \ldots | "Z"

  <letter> ::= <lower> | <upper>

  <digit> ::= "0" | "1" | \ldots | "9"

  <idchar> ::= <lower> | <upper> | <digit> | "_"

  <variable> ::= <letter> <idchar>*

  <ovariable> ::= <variable> | "_"

  <predicate> ::= <lower> <idchar>*

  <region> ::= <upper> <idchar>*

  <guardname> ::= <letter> <idchar>*

  <abinop> ::= \verb|&*&|

  <assertion> ::= <pureassn> | <predassn> | <assertion> <abinop> <assertion> \\
              | {\color{blue} <pureassn> \verb|?| <assertion> \verb|:| <assertion>}


  <eqrel> ::= \verb|=| | \verb|!=|

  <valconst> ::= \{<digit>\}

  <valbinop> ::= \verb|+| | \verb|-| | \verb|*| | {\color{blue} \verb|/|}

  <valunop> ::= \verb|-|

  <valexpr> ::= <ovariable> | <valconst> | <valexpr> <valbinop> <valexpr> \\
    | <valunop> <valexpr> \\
    | {\color{blue} \verb|{| <valexpr> \{ "," <valexpr> \} \verb|}| }

  <valrel> ::= <eqrel> | \verb|<| | \verb|<=| | \verb|>| | \verb|>=|

  <permconst> ::= "0p" | "1p"

  <permbinop> ::= \verb|$|

  <permunop> ::= \verb|~|

  <permexpr> ::= <ovariable> | <permconst> | <permexpr> <permop> <permexpr> | <permunop> <permexpr>

  <permrel> ::= <eqrel> | \verb|#|

  <pureassn> ::= "true" | "false" | \verb|!| <pureassn> \\
    | <ovariable> <eqrel> <ovariable>
    | <valexpr> <valrel> <valexpr> \\
    | <permexpr> <permrel> <permexpr>

  <expression> ::= <ovariable> | <valueexpr> | <permexpr>

  <regionassn> ::= <region> "(" <variable> "," \{ <expression> "," \} <valexpr> ")"

  <pred> ::= <predicate> [ "(" [ <expression> ] \{ "," <expression> \} ")"]

  <cell> ::= <valueexpr> "|->" <valueexpr> | <valueexpr> "|->" "#cells" "(" <valueexpr> ")"

  <guard> ::= <guardname> | <guardname> "[" <permexpr> "]" \\
  | {\color{blue} <guardname> "(" <expression> \{ "," <expression> \} ")" }

  <guards> ::= <variable> "@" <guard> | <variable> "@" "(" <guard> \{ \verb|*| <guard> \} ")"

  <predassn> ::= <regionassn> | <pred> | <cell> | <guards>

\end{grammar}

\section{Symbolic Execution}

The symbolic state will consist of: program variable assignments; pure facts; predicates; guards; and regions.
All of these can mention (logical) variables.

The program variable assignment will associate value expressions with program variables.
At the beginning of executing a procedure, all program variables used in the procedure should be associated with a logical variable with the same name.

The pure facts will be assertions about the logical variables.

The predicates will be a multiset of (predicate name, expression list) pairs.
There will be a built-in predicate called $\mathsf{\#cell}$ to represent heap cells.
There will be a table of information about predicates that describes: their number and type of parameters; pure facts that can be obtained from predicates; how/when to roll and unroll them.
For instance, $\mathsf{\#cell}$ will have two parameters of value type, and producing a cell will enforce that its address (first parameter) is distinct from the address of any other extant cell --- a pure fact.


\section{Concerning region identifiers}

Within the symbolic state, region identifiers are associated with:
\begin{itemize}
  \item Region type information;
  \item Region state information; and
  \item Guard resources.
\end{itemize}

The only interesting condition on region identifiers is their equality.
If region identifiers are equal, then the region type information and region state information must agree.
In theory, we might have information about two regions, and not know if they are actually the same region.
At some future time, we may come to identify these two regions (or discriminate them).
In addition, we may wish to introduce aliases for existing regions.

At the very least, we should have some mechanism for handling the aliasing of regions.

\section{On Producing and Consuming Assertions}

Producing an assertion means assuming that it is true and adding the appropriate resources to the symbolic state.
Consuming an assertion means asserting that it is true and removing the appropriate resources from the symbolic state.
Producing and consuming assertions happen in a variety of contexts; the main difference between the behaviour in these contexts is how variables are handled.

\paragraph{Producing a precondition.}
At the start of verifying a procedure, the precondition is produced in an initially empty symbolic state.
All assertion variables are treated as universals; that is, they are bound to fresh logical variables.
The program variables that correspond to the arguments of the procedure are bound to the logical variables in the same way as assertion variables are.

For example, suppose that $\mathtt{proc(x, w)}$ has the following precondition:
\[
  x \mapsto y \mathbin{\verb|&*&|} y \mapsto z \mathbin{\verb|&*&|} z = y
\]
This should produce the following initial symbolic state, (up to naming of variables):
\begin{itemize}
  \item Logical variables: $\mathit{vx}, \mathit{vy}, \mathit{vz}, \mathit{vw}$.
  \item Pure facts: $\mathit{vz} = \mathit{vy}$
  \item Program variables: $\mathtt{x} := \mathit{vx}$, $\mathtt{w} := \mathit{vw}$
  \item Heap chunks: $\mathit{vx} \mapsto \mathit{vy}$, $\mathit{vy} \mapsto \mathit{vz}$.
\end{itemize}
As a side-effect, the following binding context for assertion (and program) variables will be produced:
\[
  x \mapsto \mathit{vx}, y \mapsto \mathit{vy}, z \mapsto \mathit{vz}, w \mapsto \mathit{vw}
\]
This is relevant, because subsequent assertions in the code (notably the postcondition) should use these bindings, although preferring the present value of program variables in the case of collisions.

In practice, this binding context will be implicit by  binding $x$ to the logical variable $\mathtt{VIDNamed}\ x$.
Subsequent assertions will only have to look to see which logical variables in the assumptions include those with the $\mathtt{VIDNamed}$ constructor.
($\mathtt{VIDNamed}$ variables must be guaranteed not to be introduced by other means.)

\paragraph{Consuming a postcondition.}
At the end of verifying a procedure, the postcondition is consumed on the current symbolic state.
This may generate (angelic) non-determinism, as there may be some choice as to which resources and regions in the assertion map to which in the symbolic state.

Obviously, we need a variable in the postcondition to correspond to the return value of a procedure.
There are a few options:
\begin{itemize}
  \item Use a reserved variable name (e.g. ``ret'') to always mean the return value.
    This is simple, and probably easy, but maybe not the most elegant solution.
  \item Have the user specify in the ensures clause the name of the variable to bind to the return value.
    Some syntax like \texttt{ensures ret. ret = 12;} might do it.
  \item As above, but allow the user to not specify a binding if it isn't needed.
  \item (Risky compromise) allow the user to specify, but have a default.
\end{itemize}

\textbf{Decision:}
For now, I will go with ``ret'' and we can tweak later.
(Note, the binding of the return value will always have precedence.
If, for example, there is a program variable or argument called \texttt{ret}, then it cannot be referred to as ``ret'' in the postcondition.)

Another question is that of referring to program variables in the postcondition.
Again, there are a few options:
\begin{itemize}
  \item Postcondition variables do not refer to program variables.
    However, they can refer to variables mentioned in the precondition (and possibly other assertions).
    These include the original values of the arguments.
    All other named variables would be treated as existentials.
  \item Postcondition variables can refer to program variables.
    In this case, they would refer to these in preference to any other variables (apart from the return).
\end{itemize}

The first is what Verifast uses.
I think there's a good justification for it in that the program variables are technically out of scope after the return.

\paragraph{Assertions.}
How should we handle assertions in the code?
\begin{enumerate}
  \item Treat them as ``midconditions'' and use them to break up the program verification into (ideally) linear segments.
    In such a scenario, they must describe the state fully enough that the subsequent code can be verified.
    In particular, they must mention all the resources needed for the continuation.
    An advantage of this is that assertions can be used as loop invariants.
\end{enumerate}

\end{document}
